A weakly structured stem for human origins in Africa

### 2022-09-02 paper.tex
added comment on CSFS
removed comment

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
These populations evolve independently until 119ka
(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa. 
\sgcomment{Here my read is that the statement about stem 2 in Mende is shared across both models.
The statement about the Nama is specific to the merger model. Make this explicit?
}

@@

Previous studies have found support for archaic hominin admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. However, none of these studies considered a weakly
structured stem. We validated our inferred models with additional independent
approaches. We find that the observed cSFS (conditioned on the derived allele
being carried in the Neanderthal sample) is very well-described by the merger
model (Figures~\ref{fig:validation}A-C and
\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
even though this statistic was not used in the fit. Our best-fit models outperform
archaic hominin admixture models fit directly to the cSFS
(for example, compare to Figure~1 in
Durvasula and Sankararaman (2020)\citep{Durvasula2020-td}).

>>>>

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
These populations evolve independently until 119ka
(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

Previous studies have found support for archaic hominin admixture in Africa using
two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
\citep{Speidel2019-nj}. However, none of these studies considered a weakly
structured stem. We validated our inferred models with additional independent
approaches. We find that the observed cSFS (conditioned on the derived allele
being carried in the Neanderthal sample) is very well-described by the merger
model (Figures~\ref{fig:validation}A-C and
\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
even though this statistic was not used in the fit. Our best-fit models outperform
archaic hominin admixture models fit directly to the cSFS
(for example, compare to Figure~1 in
Durvasula and Sankararaman (2020)\citep{Durvasula2020-td}). 
Specifically, it is the addition of migration between stems that results in a qualitative improvement 
of the agreement 
(compare Figures \ref{fig:supp-csfs-merger-without-stem-migration}  and \ref{fig:supp-csfs-merger-with-stem-migration}).
